Draft Final Report: Measles risk assessment, modelling, and benefit--cost analyses for New Zealand

### 2015-04-24 draftfinalreport_v3.Rnw
drop figure 13

\begin{figure}
     \begin{center}
     \includegraphics[width=0.9\textwidth]{naive_map.pdf}
     \end{center}
     \caption{Numbers of na\"{\i}ve individuals per District Health Board, using national immunity data (\autoref{table:sero})}
     \label{fig:naive_map}
\end{figure}

@@

The majority of the na\"{\i}ve among the general New Zealand population are in their first years of life (\autoref{fig:naive}). However, the distribution of na\"{\i}ve at a national level shows that the recent MMR vaccination schemes are reducing the proportions of na\"{\i}ve population in the 3--5 year old age group to greater levels than in older young people  (\autoref{fig:naive}). Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of na\"{\i}ve people are in DHBs with larger urban areas (\autoref{fig:naive_map}). The distribution of the numbers of na\"{\i}ve and the total na\"{\i}ve populations per DHB, assuming national immunisation and immunity rates are representative, are given in the following figures in the appendix (\autoref{sec:app}): 
\begin{itemize}[noitemsep,nolistsep]
\item Northland: \autoref{fig:Northland}
\item Waitemata: \autoref{fig:Waitemata}
\item Auckland: \autoref{fig:Auckland}
\item Counties Manukau: \autoref{fig:Counties_Manukau}
\item Waikato: \autoref{fig:Waikato}
\item Lakes: \autoref{fig:Lakes}
\item Bay of Plenty: \autoref{fig:BayofPlenty}
\item Tairawhiti: \autoref{fig:Tairawhiti}
\item Taranaki: \autoref{fig:Taranaki}
\item Hawke's Bay: \autoref{fig:HawkesBay}
\item Whanganui: \autoref{fig:Whanganui}
\item Mid-Central: \autoref{fig:Midcentral}
\item Hutt: \autoref{fig:Hutt}
\item Capital and Coast: \autoref{fig:CapitalandCoast}
\item Wairarapa: \autoref{fig:Wairarapa}
\item Nelson Marlborough: \autoref{fig:NelsonMarlborough}
\item West Coast: \autoref{fig:WestCoast}
\item Canterbury: \autoref{fig:Canterbury}
\item South Canterbury: \autoref{fig:SouthCanterbury}
\item Southern: \autoref{fig:Southern}
\end{itemize}

@@

The distribution of the measles cases (\autoref{fig:ageandvac}) appears to reflect the distribution of na\"{\i}ve individuals in the population (\autoref{fig:naive}). The vaccination history of the cases suggests that lack of vaccination cover is the main contributor to the outbreak, though it is noticeable that a number ($>16\%$) of cases had been vaccinated at least once. However, the majority of these received only one vaccination, and were vaccinated when they were young (\autoref{fig:vaccstat}). The majority of na\"{\i}ve among the New Zealand population is clearly focused in the DHB with large urban areas (\autoref{fig:naive_map}).

@@

The results of these modelling exercises suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (\autoref{fig:r0}). However, the na\"{\i}ve population (\autoref{fig:naive} and \autoref{fig:naive_map}) and the higher $R_v$ for some of the 2009--2014 outbreaks (\autoref{fig:r0}) suggests that catch up vaccination is necessary (\autoref{table:attack}). The results of these analyses support other studies that suggest very high rates of vaccination are required to eliminate measles and prevent epidemics. Vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals was suggested to be sufficient to prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached, with approximately 11\% currently na\"{\i}ve. Analysis of measles vaccination in various regions suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics \citep{agur93, babad95, edmunds0, gay98, wallinga1}, whereas \citep{glass4} showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection making effective immunisation difficult.

@@

To estimate the benefits from additional vaccinations, as estimated from the above modelling section (\autoref{section:epidemic_modelling}), we carried out a number of analyses. Primarily, we used \autoref{eq:prop} to estimate the proportion of the na\"{\i}ve populations in each DHB (\autoref{fig:naive_map} and \autoref{table:attack}) and the national level requiring vaccination to reduce the $R_v$ to <1 (\autoref{section:epidemic_modelling}). We assume that the number of cases prevented by this is proportional to the outbreak size. However, there is a continued risk of introduction and despite $R_v$ being <1, smaller outbreaks may occur. Thus, we simulated expected outbreak sizes with $R_v$ < 1. We use these values of numbers of predicted cases prevented and numbers expected despite additional vaccination to calculate the savings.

>>>>

@@

The majority of the na\"{\i}ve among the general New Zealand population are in their first years of life (\autoref{fig:naive}). However, the distribution of na\"{\i}ve at a national level shows that the recent MMR vaccination schemes are reducing the proportions of na\"{\i}ve population in the 3--5 year old age group to greater levels than in older young people  (\autoref{fig:naive}). The distribution of the numbers of na\"{\i}ve and the total na\"{\i}ve populations per DHB, assuming national immunisation and immunity rates are representative, are given in the following figures in the appendix (\autoref{sec:app}): 
\begin{itemize}[noitemsep,nolistsep]
\item Northland: \autoref{fig:Northland}
\item Waitemata: \autoref{fig:Waitemata}
\item Auckland: \autoref{fig:Auckland}
\item Counties Manukau: \autoref{fig:Counties_Manukau}
\item Waikato: \autoref{fig:Waikato}
\item Lakes: \autoref{fig:Lakes}
\item Bay of Plenty: \autoref{fig:BayofPlenty}
\item Tairawhiti: \autoref{fig:Tairawhiti}
\item Taranaki: \autoref{fig:Taranaki}
\item Hawke's Bay: \autoref{fig:HawkesBay}
\item Whanganui: \autoref{fig:Whanganui}
\item Mid-Central: \autoref{fig:Midcentral}
\item Hutt: \autoref{fig:Hutt}
\item Capital and Coast: \autoref{fig:CapitalandCoast}
\item Wairarapa: \autoref{fig:Wairarapa}
\item Nelson Marlborough: \autoref{fig:NelsonMarlborough}
\item West Coast: \autoref{fig:WestCoast}
\item Canterbury: \autoref{fig:Canterbury}
\item South Canterbury: \autoref{fig:SouthCanterbury}
\item Southern: \autoref{fig:Southern}
\end{itemize}

@@

The distribution of the measles cases (\autoref{fig:ageandvac}) appears to reflect the distribution of na\"{\i}ve individuals in the population (\autoref{fig:naive}). The vaccination history of the cases suggests that lack of vaccination cover is the main contributor to the outbreak, though it is noticeable that a number ($>16\%$) of cases had been vaccinated at least once. However, the majority of these received only one vaccination, and were vaccinated when they were young (\autoref{fig:vaccstat}).

@@

The results of these modelling exercises suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (\autoref{fig:r0}). However, the na\"{\i}ve population (\autoref{fig:naive}) and the higher $R_v$ for some of the 2009--2014 outbreaks (\autoref{fig:r0}) suggests that catch up vaccination is necessary (\autoref{table:attack}). The results of these analyses support other studies that suggest very high rates of vaccination are required to eliminate measles and prevent epidemics. Vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals was suggested to be sufficient to prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached, with approximately 11\% currently na\"{\i}ve. Analysis of measles vaccination in various regions suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics \citep{agur93, babad95, edmunds0, gay98, wallinga1}, whereas \citep{glass4} showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection making effective immunisation difficult.

@@

To estimate the benefits from additional vaccinations, as estimated from the above modelling section (\autoref{section:epidemic_modelling}), we carried out a number of analyses. Primarily, we used \autoref{eq:prop} to estimate the proportion of the na\"{\i}ve populations in each DHB (\autoref{table:attack}) and the national level requiring vaccination to reduce the $R_v$ to <1 (\autoref{section:epidemic_modelling}). We assume that the number of cases prevented by this is proportional to the outbreak size. However, there is a continued risk of introduction and despite $R_v$ being <1, smaller outbreaks may occur. Thus, we simulated expected outbreak sizes with $R_v$ < 1. We use these values of numbers of predicted cases prevented and numbers expected despite additional vaccination to calculate the savings.
